MCC Models

### 2019-08-08 papers/full_paper/SI_appendix.Rmd
rework pragmatics model section in SI

\begin{equation} 
P_{L_1}(r, \mathcal{L}|u)\propto P_{S_1}(u|r)P( \mathcal{L})P(r)
\end{equation}

The right-hand side of this equation has three terms: the prior distribution over referents $P(r)$, the prior distribution over lexica $P(\mathcal{L})$, and the likelihood that a speaker would produce an utterance $u$ given a referent $r$ ($P_{S1}(u|r)$).
In the situation we model, the prior on referents $P(r)$ is a categorical distribution over three objects in a scene, which we posit could be non-uniform due to what is in common ground (see below for common ground manipulations). 
Because the labels produced by the speaker are all novel words, the listener has no substantive knowledge about the lexica (label--type mappings) and thus the prior over the two lexica (described above) is uniform.

The pragmatic listener updates their beliefs about both the referent and the lexicon by reasoning about the speaker, assumed to produce utterances to convey the referent to the listener by being a soft-max rational agent (with degree of rationality $\alpha$) with a utility function defined in terms of the informativity of an utterance for a referent:

\begin{equation} 
P_{S_1}(u|r)\propto exp(\alpha Informativity(u;r))
\end{equation}

As spelled out in @goodman2013knowledge, the informativity of an utterance for a referent is the (log) probability that a naive listener $L_0$ would select that referent given that utterance.

\begin{equation} 
Informativity(u; r) = \ln P_{L_0}(r|u)
\end{equation}

We assume that the speaker believes the literal listener to also be uncertain about the lexicon; that is, the pragmatic listener believes that the speaker (correctly) believes that the listener does not know the lexicon. 
Because the literal listener is assumed to not know the lexicon, the informativity of the utterance averages over the possible lexical, or words meanings, that $L_0$ considers:

\begin{equation} 
P_{L_0}(r|u) = \sum_L P_{L_0}(r, \mathcal{L}|u)
\end{equation}

where $P_{L_0}(r, \mathcal{L}|u)  \propto \mathcal{L}_{lit} P(r) P(\mathcal{L})$.
$P(r)$ again denotes the prior probability of a referent, $\mathcal{L}_{lit}$ encodes the literal meaning of an utterance given a particular lexicon, returning 1 if the utterance is literally true and 0 if the utterance is literally false. As mentioned above, because $L_1$ does not know the lexicon, the semantics of the words contained in the utterance (i.e., the labels) offer no information about the referent. The non-linguistic aspect of the utterance (the turning or pointing), however, do. As described above, the semantics of turning to one of the tables is roughly equivalent to saying "It's an object on that table". That is, $\mathcal{L}_{lit}$ returns 1 for a referent that is on the table the agent turned to and 0 for referents on the other table.

>>>>

\begin{equation} 
P_{L_1}(r, \mathcal{L}|u)\propto P_{S_1}(u|r, \mathcal{L})P( \mathcal{L})P(r)
\end{equation}

The right-hand side of this equation has three terms: the prior distribution over referents $P(r)$, the prior distribution over lexica $P(\mathcal{L})$, and the likelihood that a speaker would produce an utterance $u$ given a referent $r$ and their lexicon $\mathcal{L}$: ($P_{S1}(u|r, \mathcal{L})$).
In the situation we model, the prior on referents $P(r)$ is a categorical distribution over three objects in a scene, which we posit could be non-uniform due to what is in common ground (see below for common ground manipulations). 
Because the labels produced by the speaker are all novel words, the listener has no substantive knowledge about the lexica (label--type mappings) and thus the prior over the two lexica (described above) is uniform.

The pragmatic listener updates their beliefs about both the referent and the lexicon by reasoning about the speaker, assumed to produce utterances to convey the referent type ($r^t$) to the listener by being a soft-max rational agent (with degree of rationality $\alpha$) with a utility function defined in terms of the informativity of an utterance for a referent type (r^t):

\begin{equation} 
P_{S_1}(u^{a,l} | r^t, \mathcal{L})\propto Informativity(u^a;r^t)^\alpha \cdot P(u^l \mid \mathcal{L})
\end{equation}

The speaker's utterances are (action, label) pairs $(u^a, u^l)$ (i.e., we assume the speaker must turn to one of the locations and must produce a label). The label the speaker produces $u^l$ depends upon their lexicon $\mathcal{L}$ (i.e., they choose the label that is consistent with the object under their lexicon). The action (a point or a turn) the speaker produces $u^a$ is a function of that signal's informativity. That is, because the listener does not know the meaning of the labels, the labels carry no information to the listener.
The informativity of the utterance (action) for a referent is the probability that a naive listener $L_0$ would select a the type of referent ($r^t$) given that utterance (action)[@goodman2013knowledge].

\begin{equation} 
Informativity(u^a; r^t) = P_{L_0}(r^t|u^a)
\end{equation}

\noindent where the probability that the naive listener $L_0$ would select that a referent of a given type $r^t$ given the utterance $u^a$ is given by Bayes' Rule:

\begin{equation} 
P_{L_0}(r^t|u) = \sum_{r_i \in t} P_{L_0}(r_i | u) \propto \mathcal{L}_{point} P(r^i)
\end{equation}

\noindent where \mathcal{L}_{point} encodes the meaning of the actions (points), and is simply a truth-function stipulating the location of the referent must be at the location of the point / turn.
We assume in this model that the speaker is trying to convey the type of the referent $r^t$ rather than the individual token referents $r^i$; thus, the relevant probability for purposes of informativity is the type probability $P_{L_0}(r^t|u)$, which is simply a sum of the token probabilities for tokens that are of the same type $ \sum_{r_i \in t} P_{L_0}(r_i | u)$.

$P(r)$ again denotes the prior probability of a referent, $\mathcal{L}_{point}$ encodes the literal meaning of a point, returning 1 if the object is at the location of the point and 0 if the utterance if not. As mentioned above, because $L_1$ does not know the lexicon, the semantics of the words contained in the utterance (i.e., the labels) offer no information about the referent. The non-linguistic aspect of the utterance (the turning or pointing), however, do. As described above, the semantics of turning to one of the tables is roughly equivalent to saying "It's an object on that table".

The above formulation of a literal interpretation model based only on the semantics of the non-linguistic signals (points) can also be derived by positing a literal interpretation model that updates their beliefs according to the literal meanings of both labels and points, but which is uncertain about the meaning of the labels:  $P_{L_0}(r, \mathcal{L}|u)  \propto \mathcal{L}_{lit} P(r) P(\mathcal{L})$.
